---
title: "Appendix S3: Statistical Models, Results and Figures"
output:
  pdf_document: default
  html_document: default
---

This document lays out R code for all analyses and creates figures for "Climate and habitat continuity interact to alter contemporary dispersal potential", by: Lauren L. Sullivan, Zoe M. Portlas, Kelsey M. Jaeger, Mercedes Hoffner, and Jill A. Hamilton (submitted 2023 to *Ecology & Evolution*). Code was created by Lauren Sullivan, please contact LLSULL@msu.edu with questions. 





```{r, echo=FALSE}
rm(list=ls())


```

```{r, results = "hide", message = FALSE}
library(plyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(lsmeans)
library(tidyverse)
library(MuMIn)
library(rootSolve)
library(pracma)
library(gridExtra)
library("cowplot")

##Data
gdat <- read.csv("geum_data.csv", header=TRUE) #trait and climate data
tvdat <- read.csv("terminalvelocity_data.csv", header=TRUE) #terminal velocity data
height <- read.csv("height_data.csv", header=TRUE) #flowering height data from herbaria
wind <- read.csv("wind_data.csv", header=TRUE) #wind data

#data clean
gdat$Replicate <- as.factor(gdat$Replicate) #rep per mom

```

#Data Analysis and Figure Creation

We first analyze all diaspore trait data and create the figures in the manuscript. We examine diaspore mass, morphology, and terminal velocity.

```{r, warning=FALSE, fig.width=6}

#####
#
#  MASS MEASUREMENTS
#
#####

##DIASPORE MASS 
masstest <- lmer(sqrt(Weight_Tot) ~ Region * log(DD5_TOT) + (1|Population_Code) + 
              (1|Year), data= gdat, REML=FALSE)
summary(masstest)
r.squaredGLMM(masstest)

#Set up color palatte for all figures
theme_ba <- function (base_size = 12, base_family = "") 
{
  theme_grey(base_size = base_size, base_family = base_family) %+replace% 
  theme(axis.text = element_text(size = rel(0.8)), 
      axis.ticks = element_line(colour = "black"),
  legend.key = element_rect(colour = "grey80"), 
      panel.background = element_rect(fill = "white", 
  colour = NA), panel.border = element_rect(fill = NA, 
  colour = "black"), panel.grid.major = element_line(colour = "white", 
  size = 0.2), panel.grid.minor = element_line(colour = "white", 
  size = 0.5), strip.background = element_rect(fill = "light grey", 
  colour = "grey50", size = .75))
}

##Create Figure Panel 3a
mass <- ggplot(gdat, aes(x=log(DD5_TOT), y=sqrt(Weight_Tot), linetype = Region))+
  geom_point(aes(col=Region),size=3)+
	geom_point(shape=1,size=3,colour='black')+
	geom_smooth(aes(col=Region, fill = Region), method='lm', alpha = 1)+
	theme_ba()+
	scale_color_manual(values=c('gray20','gray70'))+
  scale_fill_manual(values=c('gray40','gray40'))+
  scale_linetype_manual(values=c("dashed", "dashed"))+
	ylab('sqrt(Total Mass (mg))')+
  theme (legend.position="none" )+ 
  labs(tag = "A)")+
  theme(axis.title.x=element_blank())


#####
#
#  MORPHOLOGY MEASUREMENTS
#
#####

##TOTAL DIASPORE LENGTH 
lengthtest <- lmer(sqrt(Dispersal_and_seed_mm) ~ Region * log(DD5_TOT) + 
          (1|Population_Code/Sample_ID) + (1|Year), data = gdat, REML=FALSE)
summary(lengthtest)
r.squaredGLMM(lengthtest)


##DIASPORE AREA 
areatest <- lmer(sqrt(SeedArea) ~ Region + log(DD5_TOT) + (1|Population_Code/Sample_ID) + 
          (1|Year), data = gdat, REML=FALSE)
#summary(areatest)
#r.squaredGLMM(areatest)


##DIASPORE SHAPE INDEX 
shapetest <- lmer(sqrt(SeedShapeIndex) ~ Region + log(DD5_TOT) + 
          (1|Population_Code/Sample_ID) + (1|Year), data = gdat, REML=FALSE)
#summary(shapetest)
#r.squaredGLMM(shapetest)


##Create Figure Panel 3b
lnth <- ggplot(gdat, aes(x=log(DD5_TOT), y=sqrt(Dispersal_and_seed_mm),
                         col=Region, linetype = Region))+
	geom_point(aes(col=Region),size=3)+
  geom_point(shape=1,size=3,colour='black')+
	geom_smooth(aes(col=Region, fill = Region),method='lm', alpha = 1)+
	theme_ba()+
	scale_color_manual(values=c('gray20','gray70'))+
  scale_fill_manual(values=c('gray40','gray40'))+
  scale_linetype_manual(values=c("dashed", "dashed"))+
	ylab('sqrt(Total Length (mm))')+
  theme (legend.position="none" )+ 
  labs(tag = "B)")+
  theme(axis.title.x=element_blank()) 


#####
#
#  TERMINAL VELOCITY MEASUREMENTS
#
#####

##TERMINAL VELOCITY 
tvtest <- lmer(sqrt(tv) ~ Region * log(DD5_TOT) + (1|Population/Sample_ID) + (1|Year), 
          data=tvdat, REML=FALSE)
summary(tvtest)
r.squaredGLMM(tvtest)


##Creates Figure Panel 3c
tv <- ggplot(tvdat, aes(x=log(DD5_TOT), y=sqrt(tv),
                        col=Region, linetype = Region))+
	geom_point(aes(col=Region),size=3)+
  geom_point(shape=1,size=3,colour='black')+
	geom_smooth(aes(col=Region, fill = Region),method='lm', alpha = 1)+
	theme_ba()+
	scale_color_manual(values=c('gray20','gray70'))+
  scale_fill_manual(values=c('gray40','gray40'))+
  scale_linetype_manual(values=c("dashed", "dashed"))+
	ylab('sqrt(Terminal Velocity (m/s))')+
  theme (legend.position="none" )+
  labs(tag = "C)")+
  theme(axis.title.x=element_blank())

#to make the legend readable.
leg <- ggplot(tvdat, aes(x=log(DD5_TOT), y=sqrt(tv),col=Region))+
	geom_point(aes(col=Region),size=3, position = position_dodge(width = 0.1))+
	scale_color_manual(values=c('gray20','gray70'))+
  theme_bw()

g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend<-g_legend(leg)

##Creates Figure 3
Fig3_a <- plot_grid(mass, lnth, tv, 
                    nrow = 1)

title_x <- ggdraw()+
                  draw_label(bquote('log(Degree Days above' ~5^o~C*')'),
                             fontface = "bold",
                             x = .35, y = .9, 
                             hjust = 0,
                             size = 16)

Fig3_b <- plot_grid(Fig3_a, NULL, title_x,
                       ncol = 1,
                       rel_heights = c(1,0.05, 0.1))

Fig3_all <- plot_grid(Fig3_b, mylegend,
                    rel_widths = c(10,1),
                    nrow = 1)

Fig3_all

# tiff("Fig3.tiff", units="in", width=8.5, height=4, res=300, compression = 'lzw')
# Fig3_all
# dev.off()
```


Next we take the terminal velocity data and translate it into a dispersal potential ability at the 99% tail of the dispersal kernel.


```{r, warning=FALSE}
## Examine height differences by region
summary(aov(height_cm ~ Region, data = height))


##Translate TV into max dispersal

height_avg <- ddply(height, .(Region), summarize, avg_flwr_ht = mean(height_cm)/100)
wind_avg <- ddply(wind, .(measurement), summarize, wind_average =
          (mean(value)*1000)/3600)  #km/h into m/s

height_avg$max_wind <- 0
height_avg$equal_wind <- 0
height_avg[2,3] <- wind_avg[2,2]      #add prairie wind - which is avg of June winds at
                                      #   Geum common garden for 2017 and 2018
height_avg[1,3] <- wind_avg[2,2]/2    #assume alvar wind is 50% of prairie wind
height_avg[,4] <- wind_avg[2,2]       #for equal wind between prairie and alvar.

tvdat_all <- merge(tvdat, height_avg, by="Region")


##Creating the kernels using the WALD model (based on Katul et al. 2005)
#Wind-based, and other non-plant parameters
kappa <- .3 	#set between .3 and .4; (.3 = canopies that are very dense)
h <- .2  		#Canopy height.
sigma2 <- kappa*h*(2*(1/2)) #sigma2 <- kappa*h*(2*(sigmaw/U)) - sigmaw/U = ratio of 1/2.
g <- 9.8 #m/s
rho <- .001225*1000000 #g/m3, density of air at 15oC at sea level is 0.001225 g/cm3

#Inverse-gaussian params (WALD), calculated from data and wind parameters
tvdat_all$lambda <- (tvdat_all$avg_flwr_ht/sqrt(sigma2))^2  #lambda
tvdat_all$mu <- (tvdat_all$avg_flwr_ht*tvdat_all$max_wind)/tvdat_all$tv   #mu

#plot exploring mean dispersal distance by region
ggplot(tvdat_all, aes(x=Region, y=mu))+
  geom_boxplot()

#Use CDF to extract tail, or the distance the farthest 1% will travel.
results_dstar <- matrix(nrow=0,ncol=2)
row <- as.vector(1:nrow(tvdat_all))
max_tail <- 10000

for(i in 1:length(row)){
  #i=3
  temp <- tvdat_all[i,]
  lambda <- temp$lambda
  mu <- temp$mu

  fun_99<- function(x) (.5*erfc((sqrt(lambda/x)*(mu-x))/(sqrt(2)*mu)) +
        .5*(exp((2*lambda)/mu))*erfc((sqrt(lambda/x)*(mu+x))/(sqrt(2)*mu))-.99)
  dstar_99 <- uniroot(fun_99, c(.00001,max_tail))$root

  new <- cbind(temp,  dstar_99)
  results_dstar <- rbind(results_dstar, new)

  #print(i/length(row))  #keeps track of progress
}

#write_csv(results_dstar, "GEUM_dispersal_kernel_integration.csv")

##DISPERSAL POTENTIAL
disp99_test <- lmer((dstar_99) ~ Region * log(DD5_TOT) + (1|Population/Sample_ID) +
          (1|Year), data=results_dstar, REML=FALSE)
summary(disp99_test)
r.squaredGLMM(disp99_test)


##Creates Figure 4
Fig4_a <- ggplot(results_dstar, aes(x=log(DD5_TOT), y=dstar_99,col=Region, 
                                    linetype = Region))+
	geom_point(aes(col=Region),size=3, position = position_dodge(width = 0.1))+
  geom_point(shape=1,size=3,colour='black', position = position_dodge(width = 0.1))+
	geom_smooth(aes(col=Region, fill = Region),method='lm', alpha = 1)+
	theme_ba()+
	scale_color_manual(values=c('gray20','gray70'))+
  scale_fill_manual(values=c('gray40','gray40'))+
  scale_linetype_manual(values=c("dashed", "dashed"))+
  xlab(bquote('log(Degree Days above' ~5^o~C*')'))+
  ylab('Dispersal Distance (m)')+  
  theme (legend.position="none" )+
  scale_y_continuous(trans = 'log2')


Fig4_all <- plot_grid(Fig4_a, mylegend,
                    rel_widths = c(3,0.5),
                    nrow = 1)
Fig4_all

# tiff("Fig4.tiff", units="in", width=5, height=4, res=300, compression = 'lzw')
# Fig4_all
# dev.off()

```
